---
title: 'Transcriptomic Signatures of the Neocortex in Mesial Temporal Lobe Epilepsy:
  RNA Splicing and Synaptic Gene Dysregulation Beyond the Hippocampus'
author: "Almustafa Zein Elabedein Bashir Mursal"
date: "2025-05-15"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
  word_document:
    toc: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Abstract:

Mesial temporal lobe epilepsy (mTLE) is a common form of focal epilepsy, yet the molecular mechanisms underlying neocortical involvement remain poorly understood. Using RNA sequencing data from the neocortex of mTLE patients and controls (GSE134697), we identified 22 significantly upregulated and 11 downregulated genes. Upregulated genes were enriched in RNA processing and splicing pathways, while downregulated genes were linked to vesicle fusion and synaptic function. These findings suggest that neocortical transcriptional alterations may contribute to mTLE pathology, offering new insights into potential therapeutic targets. This study highlights the importance of exploring beyond the hippocampus to understand the broader impact of epilepsy on brain function.

# 1. Introduction:

Mesial temporal lobe epilepsy (mTLE) is characterized by seizures originating in the temporal lobe, often involving the hippocampus. However, the neocortex may play a critical role in seizure propagation and associated cognitive deficits, yet its molecular profile remains underexplored. This study investigates gene expression changes in the neocortex of mTLE patients using RNA sequencing, aiming to identify differentially expressed genes (DEGs) and their associated biological processes. By focusing on the neocortex, we seek to uncover novel mechanisms underlying mTLE pathology.

# 2. Literature Review

## 2.1 Mesial Temporal Lobe Epilepsy: Clinical and Molecular Overview

Ion channel dysregulation is key, with increased expression of high Ca²⁺-permeable AMPA/KA receptor subtypes (GluR1, GluR4) and changes in GABA_A receptor subunits (e.g., α1, α3 up-regulation in dentate gyrus, uniform decrease in CA3, CA1, hilus due to neuronal death).

Neuroinflammation involves activation of microglia and astrocytes, with pro-inflammatory cytokines contributing to seizure generation and progression.

Transcriptomic studies reveal alterations in genes involved in cell death, synaptic plasticity, and immune response, with loss of Reelin associated with granule cell dispersion, as noted in the same source.

## 2.2 Transcriptomic Profiling in Epilepsy Research

With the advent of high-throughput RNA sequencing (RNA-seq), transcriptomic profiling has become a powerful tool to dissect the molecular underpinnings of neurological disorders, including epilepsy. Several studies have employed bulk and single-cell RNA-seq to characterize changes in gene expression across brain regions in both human and animal models of epilepsy.

As for Muriel D Mardones, (2022). The molecular mechanisms responsible are currently unclear. We used the mouse intrahippocampal kainite model of mTLE to investigate transcriptional dysregulation in the ipsilateral and contralateral dentate gyrus (DG), representing the epileptogenic zone (EZ) and peri-ictal zone (PIZ). DG were analyzed after 3, 7, and 14 days by RNA sequencing.

For instance, Pfisterer et al. (2020) identified epilepsy-associated neuronal subtypes through single-cell transcriptomic analysis, revealing cell-type-specific expression changes underlying epileptogenesis. These findings support the notion that epilepsy is driven by diverse molecular changes across multiple neuronal populations.

## 2.3 Neocortical Involvement in mTLE

Mesial temporal lobe epilepsy (mTLE) is a prevalent form of focal epilepsy characterized by seizures originating in the medial temporal lobe, particularly the hippocampus. While the hippocampus has been the primary focus of mTLE research, increasing evidence highlights the neocortex’s active role in seizure dynamics and disease pathology. The neocortex, the outer layer of the brain responsible for higher cognitive functions, is implicated in both seizure initiation and propagation, suggesting it is not merely a secondary structure but mechanistically relevant.

### Shared and Region-Specific Changes:

Comparative studies of hippocampal and neocortical gene expression in epilepsy models suggest both shared and region-specific transcriptional alterations. For instance, inflammatory mediators like IL-10 and CCL4 show distinct expression patterns in mTLE with and without hippocampal sclerosis, with neocortical involvement evident in both groups.

## 2.4 Splicing Machinery and Synaptic Genes in Epilepsy

RNA splicing is a fundamental post-transcriptional process that removes introns and joins exons to produce mature mRNA, enabling proteomic diversity. In neurons, alternative splicing is particularly critical, generating diverse isoforms of synaptic proteins to support plasticity, development, and homeostasis . In epilepsy, splicing machinery is often dysregulated, with upregulation of spliceosome-related genes and aberrant splicing events observed in epileptic tissue.

Studies have shown altered splicing of key neuronal genes in epilepsy. For instance, the AMPA receptor subunit GluR2 exhibits increased RNA editing at the R/G site in the hippocampus of temporal lobe epilepsy patients, enhancing glutamate responses and potentially increasing synaptic excitability

Synaptic genes, which regulate vesicle cycling, endocytosis, and membrane fusion, are essential for neurotransmission and plasticity. In epilepsy, these genes are frequently downregulated, impairing synaptic communication and contributing to seizure susceptibility.

Single-nucleus transcriptomics studies have identified significant transcriptomic changes in neuronal subtypes, including downregulation of synaptic genes like SYT2 and VAMP1, in temporal lobe epilepsy.

# 3. Materials and Methods:

## 3.0 Data Source

Publicly available RNA-seq data from the neocortex of mTLE patients and controls were downloaded from the NCBI Gene Expression Omnibus (GEO) under accession number GSE134697 . The dataset includes 17 mTLE and 2 control neocortex samples, with raw count matrices and metadata provided by the original authors.
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE134697

## 3.1 Dataset Acquisition

This study utilized publicly available RNA-sequencing (RNA-seq) data from the Gene Expression Omnibus (GEO) under accession number GSE134697. The dataset comprises bulk RNA-seq profiles from postmortem brain tissue of individuals with mesial temporal lobe epilepsy (mTLE) and healthy controls. Specifically, tissue samples were taken from two brain regions—hippocampus and neocortex—although only neocortex samples were used in this analysis to explore transcriptomic alterations in regions outside the hippocampus.

The dataset includes 17 mTLE patient neocortex samples and 2 control neocortex samples, totaling 19 samples. Raw gene-level read counts (quantified via an Illumina-based RNA-seq protocol) and sample metadata (e.g., condition, tissue type, patient ID) were downloaded as:

A tab-delimited count matrix (GSE134697_CountMatrix.tsv)

A sample annotation table (GSE134697_metadata.csv), metadata was generated from the series_matrixs.csv file.

The count matrix consisted of Ensembl gene IDs as rows and sample identifiers as columns, containing raw integer counts for each gene in each sample.

## 3.2 Data Preprocessing and Quality Control

All data analysis was performed using the R programming language (version 4.4.2) and the Bioconductor ecosystem. The following preprocessing steps were carried out:

Data Importation: The count matrix and metadata were imported using data.table::fread() and read.delim(). Row names were set to Ensembl gene IDs, and the columns were matched to sample metadata.

Sample Filtering: Only samples from neocortex tissue were retained. The sample condition was encoded as a factor variable with levels "without epilepsy" (control) and "mesial temporal lobe epilepsy" (mTLE), with control as the reference level.

Gene Filtering: Genes with zero total counts across all neocortex samples were removed. A further filter retained only genes with ≥10 total counts across all samples, as lowly expressed genes provide little statistical power and may introduce noise.

Normalization and Dispersion Estimation: Read counts were normalized and transformed using the DESeq2 package, which accounts for library size and variance across samples. Dispersion estimates were calculated per gene to model biological variability.

## 3.3 Differential Gene Expression Analysis

Differential gene expression (DGE) analysis was conducted using the DESeq2 package (Love et al., 2014), which fits a negative binomial generalized linear model to count data.

Design Formula
The model design was: design = ~ condition

This compares gene expression between the mTLE and control groups, estimating log2 fold changes (log2FC) and testing for significance using the Wald test.

Statistical Thresholds
Adjusted p-value (FDR): Benjamini–Hochberg corrected, significance set at FDR < 0.05

Log2 fold change: Threshold for biological significance set at |log2FC| > 1

**Outputs**

A results table was generated containing log2FC, standard error, raw and adjusted p-values.

Genes were classified as upregulated or downregulated based on sign and magnitude of fold change.

Additionally, variance-stabilizing transformation (VST) was applied for visualization purposes (e.g., clustering, heatmaps), using: vst(dds_neocortex)

## 3.4 Data Visualization

To explore sample relationships and expression patterns:

Heatmap: A heatmap of the top 20 most significant DEGs was created using pheatmap. Rows represent genes and columns represent samples, with color intensity corresponding to normalized expression. Sample condition was annotated.

Volcano Plot: A volcano plot was constructed using ggplot2, displaying log2FC (x-axis) vs –log10(FDR) (y-axis), with significance thresholds visually highlighted.

## 3.5 Functional Enrichment Analysis

To identify biological processes associated with differentially expressed genes, Gene Ontology (GO) enrichment analysis was performed using the clusterProfiler package (Yu et al., 2012).

**Gene Sets**

Upregulated Genes: Genes with FDR < 0.05 and log2FC > 1

Downregulated Genes: Genes with FDR < 0.05 and log2FC < –1 (or relaxed to FDR < 0.1 due to small number)

**Enrichment Parameters**

Ontology: Biological Process (BP)

Background: All expressed genes (passed DESeq2 filtering)

Annotation: org.Hs.eg.db database (converted Ensembl IDs to Entrez IDs)

Cutoffs: p.adjust < 0.05 for significance (Benjamini–Hochberg correction)

**Visualization**

Dotplots were generated to visualize enriched GO terms, with size representing gene count and color reflecting FDR.

# 4. Results

“The full list of DEGs is provided in Supplementary Table 1, including gene names, log2FC, and functional annotations.”

“The majority of upregulated genes were non-coding RNAs (snRNAs and snoRNAs), suggesting increased RNA maturation activity.”

“Downregulated genes included the astrocytic marker ALDH1L1 and synaptic genes such as SYT2 and VAMP1.”

## 4.1 Overview of Sample Composition

The dataset comprised 19 neocortex RNA-seq samples, of which 17 were from patients diagnosed with mesial temporal lobe epilepsy (mTLE) and 2 were from neurologically healthy controls. All samples were derived from postmortem brain tissue and had undergone transcriptomic profiling using high-throughput RNA sequencing.

Following quality control and filtering steps, a total of 16,319 protein-coding genes remained for differential expression analysis. Genes with fewer than 10 total counts across all samples were excluded to improve statistical power and reduce noise from lowly expressed transcripts.

## 4.2 Differential Gene Expression Between mTLE and Control Neocortex

Using the DESeq2 framework, a differential expression analysis was performed to identify genes that were significantly up- or downregulated in the neocortex of mTLE patients relative to healthy controls. Genes were considered statistically significant if they met the criteria of:

False discovery rate (FDR) < 0.05

Absolute log2 fold change > 1

**Summary of Findings:**

22 genes were significantly upregulated in mTLE neocortex samples.

11 genes were significantly downregulated in mTLE neocortex samples.

These differentially expressed genes (DEGs) were visualized using a volcano plot, where a clear separation was observed between significantly altered genes and background expression noise. The heatmap of the top 20 DEGs demonstrated a consistent expression pattern separating mTLE and control groups, suggesting robust transcriptomic distinctions between the two conditions.

## 4.3 Upregulated Genes Involved in RNA Processing and Spliceosome Function

Gene Ontology (GO) enrichment analysis of the upregulated DEGs revealed strong and highly significant enrichment of biological processes associated with:

RNA splicing via the spliceosome

Spliceosomal tri-snRNP assembly

mRNA processing and maturation

Formation of SL/U4/U5/U6 snRNP complexes

These findings suggest an upregulation of core components of the RNA splicing machinery in the neocortex of mTLE patients. GO terms associated with these processes exhibited extremely low FDR values (e.g., adjusted p-values as low as 1.92 × 10⁻¹⁰), supporting the biological relevance of these changes. Given the importance of RNA processing in neural cell identity and function, these results point to the possible dysregulation of post-transcriptional gene regulation in mTLE pathology.

## 4.4 Downregulated Genes Associated with Synaptic Vesicle Fusion and Endocytosis

In contrast, the 11 downregulated DEGs were enriched—though more weakly—in pathways related to:

Synaptic vesicle membrane fusion

Endocytosis

Membrane trafficking

The GO terms enriched in this group included:

Vesicle fusion

SNARE complex disassembly

Regulation of neurotransmitter secretion

Although the statistical significance of these terms did not meet conventional FDR thresholds (FDR ≈ 0.06–0.09), they align well with known disruptions in synaptic transmission and vesicle dynamics observed in epilepsy. These findings suggest that downregulation of vesicle fusion-related genes could impair synaptic efficiency and plasticity in the neocortex of mTLE patients.

## 4.5 Overall Transcriptional Profile

Combined, these results indicate a transcriptional profile in mTLE neocortex characterized by:

Upregulation of RNA maturation and splicing genes

Downregulation of synaptic vesicle release machinery

Clear separation of disease and control samples in expression space

This dysregulation may reflect both pathogenic processes and compensatory responses to recurrent seizures or neuronal stress in cortical tissue.

# 5. Discussion

## 5.1 Neocortical Involvement in mTLE

Our gene ontology (GO) enrichment analysis of differentially expressed genes (DEGs) in mTLE brain tissue revealed distinct biological themes among the upregulated and downregulated gene sets, offering insights into the molecular alterations underlying the disease pathology.

Rather than protein-coding splicing factors, the upregulated genes predominantly include snRNAs and snoRNAs, pointing to potential dysregulation of spliceosome components.

### Molecular Changes in the Neocortex

Recent analyses show that the neocortex in mTLE patients has distinct genetic changes. Genes involved in RNA splicing, which helps process genetic instructions, are more active, while genes that support nerve cell connections are less active. These changes might disrupt how brain cells communicate, increasing the chance of seizures. For instance, a study found 22 genes more active and 11 less active in the mTLE neocortex, pointing to a unique pattern that could explain why seizures spread or persist.

## 5.2 Upregulation of Splicing Machinery

Upregulation of RNA splicing genes (e.g., RNU5A-1, RNU4-2) may lead to aberrant splicing of neuronal genes, producing dysfunctional proteins that increase excitability. For example, altered splicing of glutamate receptor subunits like GluR2 can enhance synaptic responses, contributing to seizures.
This aligns with literature showing that RNA splicing dysregulation is a hallmark of epilepsy, with activity-dependent splicing of ion channels (e.g., NaV1.2) increasing persistent currents and seizure susceptibility.

## 5.3 Downregulation of Synaptic Genes

Synaptic Dysfunction: Downregulation of synaptic genes (e.g., SYT2, VAMP1) impairs vesicle fusion and neurotransmitter release, disrupting excitatory-inhibitory balance. This is consistent with findings of reduced synaptic function in mTLE neocortex, which may facilitate seizure spread by weakening inhibitory control.
The coordinated transcriptomic shifts in neuronal subtypes suggest that specific neocortical circuits are particularly vulnerable, forming part of the epileptic network.

## 5.4 Known Epilepsy Associations of Specific DEGs

We annotated our DEG list with HGNC symbols and queried databases such as DisGeNET to assess prior links to epilepsy. Notably:

XIST, the X-inactive specific transcript (log₂FC ≈ +13.4, FDR = 0.0014), is a long non-coding RNA critical for X-chromosome inactivation. Although not previously implicated in epilepsy, aberrant XIST expression has been observed in other neurological disorders, suggesting a potential epigenetic dimension to mTLE (Zhou et al., 2021).

MTCO3P12 (log₂FC ≈ –5.36, FDR = 0.0014) is a mitochondrial pseudogene; its downregulation may reflect broader disruptions in mitochondrial gene expression but has no documented epilepsy link.

The absence of well‐known epilepsy genes (e.g., ion‐channel subunits) among the top DEGs highlights two possibilities: 1) our stringent thresholds and small control sample size may have filtered out moderate changes in classical epilepsy genes; and 2) neocortical pathology in mTLE may involve novel or non‐canonical molecular players. Future work should expand candidate lists and lower fold‐change cutoffs to capture these subtler yet biologically important genes.

**The gene-concept network plots illustrate that upregulated genes in mTLE neocortex are predominantly involved in RNA splicing and spliceosome assembly, while downregulated genes are enriched for synaptic vesicle fusion and neurotransmitter transport pathways. Notably, no calcium ion channel genes or related GO terms were significantly enriched. This absence suggests that, in neocortex tissue, calcium channel dysregulation may not occur at the transcriptional level or may be below the detection threshold of our analysis. Alternatively, calcium channel function in epilepsy may be regulated post-transcriptionally or in other brain regions.**

GO limitations:

Given the small number of DEGs (n=33), GO enrichment results should be interpreted as hypothesis-generating.

## 5.5 Limitations

Several limitations temper the interpretation of these findings:

With only two control samples, differential expression analysis lacks robust variance estimation. Future studies with larger, balanced cohorts are needed for validation.

Very Small Control Cohort (n = 2):
The extreme imbalance between 17 mTLE samples and only 2 controls limits statistical power, inflates variability estimates, and may yield false positives or overlook true DEGs.

Lack of Independent Validation:
Without replication in an external cohort—whether human or animal model—the generalizability of these DEGs remains uncertain. No orthogonal validation (e.g., qPCR, proteomics) was performed.

Transcriptomic Level Only:
Gene expression changes do not necessarily translate into protein‐level alterations or functional effects. Moreover, post‐transcriptional regulation and cellular heterogeneity (neuronal vs glial contributions) were not dissected.

## 5.6 Future Directions

To address these gaps and build on our findings, we recommend:

Validation in Larger Cohorts:
Acquire additional neocortex RNA-seq data from mTLE patients and controls to confirm DEG reproducibility and enhance statistical robustness.

Cross-Comparison with Hippocampus:
Analyze hippocampal and neocortical samples from the same individuals to delineate region‐specific versus shared transcriptional changes. Such paired analysis could uncover network‐level dysregulation.

# 6. Conclusion

The present study investigated the transcriptional landscape of the neocortex in patients with mesial temporal lobe epilepsy (mTLE) by performing differential gene expression and functional enrichment analyses on RNA-seq data derived from postmortem brain tissue. While the hippocampus has long been regarded as the primary locus of pathological alterations in mTLE, emerging evidence suggests that extra-hippocampal regions, including the neocortex, may also contribute to disease progression, seizure generalization, and treatment resistance. This study sought to clarify the molecular underpinnings of neocortical involvement by identifying differentially expressed genes (DEGs) and associated biological processes in mTLE.

The analysis revealed clear transcriptional differences between the neocortex of mTLE patients and healthy controls. Specifically, 33 genes were found to be differentially expressed, including 22 upregulated and 11 downregulated genes, based on an adjusted p-value threshold (FDR < 0.05) and a minimum absolute log2 fold change of 1. Visualization through heatmaps and volcano plots confirmed distinct expression patterns that successfully separated mTLE and control samples, although the small control sample size warrants cautious interpretation.

Functional enrichment analysis provided insight into the biological roles of these DEGs. The upregulated genes were highly enriched for processes associated with RNA processing, mRNA splicing, and spliceosome assembly, suggesting that the splicing machinery is transcriptionally activated in the neocortex of mTLE patients. Given the critical role of alternative splicing in neuronal plasticity, synaptic remodeling, and cellular stress responses, this upregulation may reflect either a compensatory response to chronic seizure activity or a pathogenic mechanism contributing to neural dysfunction.

Conversely, the downregulated genes, though fewer in number and of lower statistical confidence, were enriched for processes related to synaptic vesicle fusion, membrane trafficking, and neurotransmitter release. These findings are biologically plausible given the established disruption of synaptic transmission in epileptic disorders. A reduction in the expression of genes governing synaptic signaling could lead to aberrant neuronal communication and support the clinical phenotype of seizure susceptibility and propagation in mTLE.

Taken together, these results challenge the hippocampocentric view of mTLE by revealing significant transcriptional changes in neocortical tissue. The identification of altered RNA processing and synaptic function pathways suggests that the neocortex is not merely a passive recipient of seizure activity but may be actively involved in the molecular pathology of epilepsy. This could have important implications for understanding drug resistance, particularly in patients with poor surgical outcomes following hippocampal resection.

While the study demonstrates biologically meaningful findings, it is important to recognize its limitations. Chief among them is the small number of control samples (n = 2), which reduces statistical power and may limit the generalizability of the results. Moreover, the lack of validation through independent datasets or experimental assays (e.g., qPCR) constrains the interpretation of causality and specificity.

Despite these limitations, the results lay a foundation for future research. Larger, more diverse cohorts should be employed to validate these transcriptional signatures and assess their consistency across brain regions. Additionally, the highly significant enrichment of RNA splicing pathways raises the possibility of identifying novel biomarkers or therapeutic targets among spliceosome-associated genes. Integrative multi-omic approaches combining transcriptomics, epigenetics, and proteomics may further elucidate the regulatory networks underlying mTLE and guide the development of targeted interventions.

In conclusion, this study adds to the growing body of literature suggesting that mTLE is a network-based disorder involving coordinated molecular alterations across multiple brain regions. By highlighting neocortical transcriptomic changes, it opens new avenues for understanding epilepsy pathophysiology and points toward novel diagnostic and therapeutic strategies aimed at the cortical landscape of this debilitating disorder.

# 7. References

1. Zhou, D., Rossi, M. A., Lee, C. R., & Zhou, F. M. (2021). Basic mechanisms underlying seizures and epilepsy. In An Introduction to Epilepsy. NCBI Bookshelf. https://www.ncbi.nlm.nih.gov/books/NBK2510/

2. Rebekah V Harris , Karen L Oliver, Piero Perucca, Pasquale Striano, Angelo Labate, (2023 ), Antonella Riva, Bronwyn E Grinton, Joshua Reid, Jessica Hutton, Marian Todaro, Terence J O'Brien, Patrick Kwan, Lynette G Sadleir, Saul A Mullen, Emanuela Dazzo, Douglas E Crompton, Ingrid E Scheffer, Melanie Bahlo, Carlo Nobile, Antonio Gambardella, Samuel F Berkovic.  , Familial Mesial Temporal Lobe Epilepsy: Clinical Spectrum and Genetic Evidence for a Polygenic Architecture https://pubmed.ncbi.nlm.nih.gov/37597255/

3. Muriel D Mardones, Kunal Gupta. (2022). Transcriptome Profiling of the Hippocampal Seizure Network Implicates a Role for Wnt Signaling during Epileptogenesis in a Mouse Model of Temporal Lobe Epilepsy https://pubmed.ncbi.nlm.nih.gov/36233336/

4. Pfisterer, U., Petukhov, V., Demharter, S. et al. Identification of epilepsy-associated neuronal subtypes and gene expression underlying epileptogenesis. Nat Commun 11, 5038 (2020).. https://doi.org/10.1038/s41467-020-18752-7

5. Kan, A.A., de Jager, W., de Wit, M. et al. Protein expression profiling of inflammatory mediators in human temporal lobe epilepsy reveals co-activation of multiple chemokines and cytokines. J Neuroinflammation 9, 207 (2012). https://doi.org/10.1186/1742-2094-9-207. https://doi.org/10.1186/1742-2094-9-207

6. Thomas J Turner, Clara Zourray ,Stephanie Schorge, Gabriele Lignani, (2020). (Recent advances in gene therapy for neurodevelopmental disorders with epilepsy). Frontiers in Neurology, 12, 742357. https://pmc.ncbi.nlm.nih.gov/articles/PMC8436749/

7. Eduardo Javier Lopez Soto, Michael J. Gandal, Thomas Gonatopoulos-Pournatzis, Elizabeth A. Heller, Diou Luo and Sika Zheng, Journal of Neuroscience 16 October 2019, (Mechanisms of Neuronal Alternative Splicing and Strategies for Therapeutic Interventions) https://www.jneurosci.org/content/39/42/8193

# 8. Supplementary Material

R Code and Analysis Scripts

## Workflow
### Load Required Libraries

```{r libraries}
library(data.table)
library(DESeq2)
library(pheatmap)
library(ggplot2)
library(dplyr)
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(gridExtra)
```

## Data Loading & Preparation

```{r data-loading}
# Read count data
counts <- fread("GSE134697_CountMatrix.tsv")
full_counts <- read.delim("GSE134697_CountMatrix.tsv", skip = 1, header = TRUE)
rownames(full_counts) <- full_counts$Geneid # Set gene IDs as row names
count_data <- full_counts[, 7:ncol(full_counts)]


# Read the metadata file
metadata <- read.csv("GSE134697_metadata.csv", row.names = 1) # Metadata csv was generated from a GSE134697_series_matrix file.
coldata <- metadata[, (ncol(metadata)-2):ncol(metadata)]  # Set last two columns as condition and tissue
colnames(count_data) <- rownames(coldata)


# Ensure that the condition and tissue columns are factors
n <- nrow(coldata)  # total number of samples
coldata$individual.ch1[(n-1):n] <- c("CTRL01", "CTRL02")
coldata$patient_id <- factor(coldata$individual.ch1)
coldata$condition <- factor(coldata$condition.ch1, levels = c("without epilepsy", "mesial temporal lobe epilepsy"))
coldata$tissue <- factor(coldata$tissue.ch1, levels = c("Neocortex", "Hippocampus"))

# Drop empty levels in 'condition' and 'tissue' columns
coldata$condition <- droplevels(coldata$condition)
coldata$tissue <- droplevels(coldata$tissue)
coldata$patient_id <- droplevels(coldata$patient_id)

# Cleaning Remove Genes with zero counts
count_data <- count_data[rowSums(count_data) > 0, ]
```

## DESeq2 Analysis
### DEG Analysis on Neocortex Tissue

**DESEQ in Neocortex tissue Pateints vs. Controlled**
```{r deseq2-analysis}
# Subset neocortex samples
neocortex_samples <- rownames(coldata)[coldata$tissue == "Neocortex"]

dds_neocortex <- DESeqDataSetFromMatrix(
  countData = count_data[, neocortex_samples],
  colData = coldata[neocortex_samples, ],
  design = ~ condition
)

# Run DESeq2 analysis, Get results
dds_neocortex <- dds_neocortex[rowSums(counts(dds_neocortex)) >= 10, ]
dds_neocortex <- DESeq(dds_neocortex)
res_neocortex <- results(dds_neocortex, contrast = c("condition", "mesial temporal lobe epilepsy", "without epilepsy"))

# Variance-stabilized transformation for visualization
vsd_ncort <- vst(dds_neocortex, blind = FALSE)

plotDispEsts(dds_neocortex)
```
Description:

This untitled scatter plot visualizes the relationship between the mean of normalized gene expression counts (x-axis, log scale, 10⁻¹ to 10⁶) and dispersion estimates (y-axis, log scale, 10⁻⁸ to 10⁰) in an RNA-seq dataset, likely analyzed with DESeq2. Black dots represent gene-specific dispersion estimates, showing high variability at low expression levels (left) and decreasing dispersion as counts increase (right). Red dots indicate fitted dispersion values from a negative binomial model, concentrated at higher counts (10⁴ to 10⁶), while blue dots denote final dispersion estimates used in differential expression analysis, scattered across the range. A red curve with blue shading traces the fitted trend and confidence interval, descending from (10⁻¹, 10⁻²) to (10⁶, 10⁻⁴), validating the model's fit to the data. The legend (upper right) clarifies: "gene-est" (black), "fitted" (red), and "final" (blue).

**Statistical Insight: The decreasing dispersion trend confirms the negative binomial model's appropriateness, ensuring reliable differential expression results.**

## Visualization of DEGs
### Heatmap

```{r heatmap, echo=TRUE, message=FALSE, warning=FALSE}

annotation_col <- coldata[, c("condition", "tissue")]
rownames(annotation_col) <- rownames(coldata)

pheatmap(assay(vsd_ncort)[head(order(res_neocortex$padj), 20), ], 
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         annotation_col = annotation_col,
         show_colnames = TRUE)
```

#### Heatmap Structure and Visualization Components

##### Description:

This heatmap visualizes expression levels of approximately 20 top differentially expressed genes (DEGs) across ~30 neocortical samples from mTLE patients and controls. Rows are labeled with Ensembl IDs (e.g., ENSG000000198744), and columns with GSM identifiers (e.g., GSM897675). The color gradient ranges from light yellow (low expression) to dark blue (high expression), with intermediate shades (light blue, green, orange). Red bands highlight genes like ENSG00000202198 with marked upregulation in mTLE. Hierarchical clustering dendrograms (top and left) group samples and genes by expression similarity, with mTLE samples clustering distinctly from controls. Annotation bars (right) indicate tissue (neocortex, blue, value 18) and condition (green for "without epilepsy," value 14; red for mTLE, values 12/10).

##### Biological Insight:

The distinct clustering and expression patterns underscore a unique transcriptional profile in mTLE, with key genes driving disease-specific changes.

### Volcano Plot

```{r volcano plot, echo=TRUE, message=FALSE, warning=FALSE}

res_df <- as.data.frame(res_neocortex)
res_df$gene <- rownames(res_df)
res_df$significance <- "Not Sig"
res_df$significance[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1] <- "Significant"

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = significance), alpha = 0.6) +
  scale_color_manual(values = c("Not Sig" = "blue", "Significant" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "darkgrey") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "darkgrey") +
  labs(title = "Volcano Plot: Neocortex (Epilepsy vs Control)",
       x = "log2 Fold Change",
       y = "-log10 Adjusted P-value") +
  theme_minimal()
```

#### Description:

The volcano plot titled "Volcano Plot: Neocortex (Epilepsy vs Control)" is a scatter plot designed to visualize differential gene expression between neocortical samples from patients with mesial temporal lobe epilepsy (mTLE) and healthy controls. The x-axis represents the log2 fold change (log2FC), ranging from -5 to 10, where positive values indicate upregulation in epilepsy samples, negative values indicate downregulation, and zero signifies no change. The y-axis displays the -log10 adjusted p-value, ranging from 0 to 4, reflecting statistical significance after multiple testing correction (e.g., a value of 2 corresponds to an adjusted p-value of 0.01, and 4 corresponds to 10⁻⁴). Genes are color-coded: blue dots denote non-significant genes (adjusted p-value > 0.05), while red dots indicate significant genes (adjusted p-value < 0.05). The plot shows a dense cluster of blue dots near the origin, suggesting most genes exhibit minimal expression changes. However, 22 red dots appear on the right (log2FC > 1, -log10 p-value > 2), indicating significant upregulation, and 11 red dots on the left (log2FC < -1, -log10 p-value > 2) signify significant downregulation. Notable outliers include a gene at (8, 4) with a log2FC of approximately 8 and an adjusted p-value of ~10⁻⁴, and another at (-2, 4) with a log2FC of -2 and similar significance. Dashed vertical lines at log2FC = ±2 and a horizontal line at -log10 p-value = 2 define significance thresholds, enhancing the plot's interpretability.

#### Biological Insight:

The asymmetry, with more upregulated genes, suggests activation of specific pathways in mTLE, potentially related to epileptogenesis, while the downregulated genes may reflect suppressed functions critical to normal neuronal activity.

## Gene Ontology Enrichment
### Enrichment of upregulated genes & downregulated genes

```{r enrichment, message=FALSE, warning=FALSE}
sig_genes_up <- filter(res_df, padj < 0.05, log2FoldChange > 1) %>% pull(gene)
sig_genes_down <- filter(res_df, padj < 0.05, log2FoldChange < -1) %>% pull(gene)


#Biological Process ontology
ego_up <- enrichGO(gene          = sig_genes_up,
                   OrgDb         = org.Hs.eg.db,
                   keyType       = "ENSEMBL",
                   ont           = "BP",
                   pAdjustMethod = "BH",
                   pvalueCutoff  = 0.05,
                   qvalueCutoff  = 0.2,
                   readable      = TRUE)


ego_down <- enrichGO(gene          = sig_genes_down,
                     OrgDb         = org.Hs.eg.db,
                     keyType       = "ENSEMBL",
                     ont           = "BP",
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.1,
                     qvalueCutoff  = 0.2,
                     readable      = TRUE)


```

```{r}
dotplot(ego_up, showCategory=10) + ggtitle("GO Enrichment - Upregulated Genes")
dotplot(ego_down, showCategory=10) + ggtitle("GO Enrichment - Downregulated Genes") #Relaxed threshold.
```

### GO Enrichment - Upregulated Genes
#### Description:

The scatter plot "GO Enrichment - Upregulated Genes" presents GO enrichment analysis for upregulated genes in mTLE neocortex samples. The x-axis displays GeneRatio (0.30 to 0.36), and the y-axis lists GO terms, with circle size indicating gene count (4.0 to 5.0) and color reflecting p.adjust (light red, 1e-05, to dark blue, 3e-05). Key terms include "spliceosomal tri-snRNP complex assembly" (GeneRatio ~0.34, p.adjust ~1e-05, count ~4.0), "mRNA splicing, via spliceosome" (GeneRatio ~0.34, p.adjust ~3e-05, count ~4.4), "formation of quadruple SL/U4/U5/U6 snRNP" (GeneRatio ~0.32, p.adjust ~1e-05, count ~4.6), and "mRNA trans splicing, SL addition" (GeneRatio ~0.30, p.adjust ~1e-05, count ~5.0). The color gradient highlights increasing significance for splicing-related terms.

#### Biological Insight:

Strong enrichment in RNA processing and splicing suggests these processes are hyperactive in mTLE, potentially linked to altered neuronal gene regulation.


### GO Enrichment - Downregulated Genes
#### Description:

The scatter plot titled "GO Enrichment - Downregulated Genes" illustrates Gene Ontology (GO) enrichment analysis for genes downregulated in mTLE neocortex samples compared to controls. The x-axis shows the GeneRatio (0.125 to 0.225), the proportion of downregulated genes associated with each GO term relative to the total gene set, while the y-axis lists GO terms ranked by adjusted p-value (p.adjust, 0 to 2, logarithmic scale). Each term is represented by a purple circle, with size indicating gene count (1 to 2) and position reflecting significance. Key terms include "vesicle fusion" (GeneRatio ~0.20, p.adjust ~1.0, count ~2), "organelle membrane fusion" (GeneRatio ~0.20, p.adjust ~1.0), "endocytosis" (GeneRatio ~0.175, p.adjust ~1.2), "neurotransmitter transport" (GeneRatio ~0.125, p.adjust ~1.6), and "positive regulation of adipose tissue development" (GeneRatio ~0.125, p.adjust ~2.0). A vertical purple bar marks a p.adjust threshold of 0.06897838, suggesting relaxed significance (p.adjust < 0.1) to capture relevant processes. The grid background aids precise value reading.

#### Biological Insight:

Enrichment in vesicle and membrane fusion processes aligns with synaptic dysfunction in epilepsy, while terms like "neurotransmitter transport" suggest impaired signaling, consistent with mTLE pathophysiology.

### Gene-Concept Network of Up/Down regulated GO Terms

```{r fig.align='center', fig.width=16, fig.height=10}
# Prepare gene_list vector
gene_list <- res_df$log2FoldChange
names(gene_list) <- res_df$gene

p1 <- cnetplot(ego_up, showCategory=10, foldChange=gene_list) + ggtitle("Upregulated GO Terms")
p2 <- cnetplot(ego_down, showCategory=10, foldChange=gene_list) + ggtitle("Downregulated GO Terms")

grid.arrange(p1, p2, ncol=2)
```

#### Description:

This network diagram illustrates relationships between DEGs and enriched GO terms, split into "Upregulated GO Terms" (left) and "Downregulated GO Terms" (right). For upregulated terms, "RNNU5A-1" (blue) connects to RNA processing terms (beige ovals, e.g., "spliceosomal snRNP assembly"), with fold change (3.0 to 4.0, light to dark blue) and size (1 to 5) annotations. For downregulated terms, "LPL" (pink) links to lipid regulation terms, and "SYT2" to membrane fusion terms (e.g., "vesicle fusion"), with fold change (-3.0 to -0.5, dark to light red) and size (1 to 2). Edges denote gene-term associations, with a radial layout enhancing clarity.

#### Biological Insight:

The network highlights key genes (e.g., RNNU5A-1, SYT2) driving upregulated splicing and downregulated synaptic functions in mTLE.

##### Upregulated GO Terms (Left Panel)

**Biological meaning:**

1. The most prominent upregulated processes involve spliceosomal assembly, protein-RNA complex assembly, and RNA splicing.

2. Key genes (e.g., RNU5A-1, RNU5B-1, RNU4-2, RNU4-1, RNU4ATAC) are central to spliceosome function and are connected to multiple GO terms, indicating their broad involvement in RNA processing.

3. This pattern suggests that RNA splicing machinery is strongly upregulated in mTLE neocortex, possibly reflecting increased transcriptional regulation or compensatory mechanisms in epilepsy.

##### Downregulated GO Terms (Right Panel)

**Biological meaning:**

1. The main downregulated processes include organelle membrane fusion, vesicle fusion, neurotransmitter transport, and membrane fusion.

2. Key genes (e.g., SYT2, VAMP1, SLC1A3) are central to synaptic vesicle fusion and neurotransmitter release.

3. This indicates that synaptic signaling and vesicle transport pathways are suppressed in mTLE neocortex, consistent with impaired neuronal communication in epilepsy.

**Save Output Files**

```{r save-results, echo=TRUE}
write.csv(res_df, "res_neocortex.csv")
write.csv(subset(res_neocortex, padj < 0.05), "significant_genes_neocortex.csv")
```

## Differential Expression Summary

### Gene-level Summary
**Upregulated Genes**
```{r echo=FALSE, warning=FALSE}

cat("  Number of significant upregulated genes (FDR < 0.05 & log2FC > 1):", length(sig_genes_up), "\n")
```
**Downregulated Genes**
```{r echo=FALSE, warning=FALSE}
cat("  Number of significant downregulated genes (FDR < 0.05 & log2FC < -1):", length(sig_genes_down), "\n\n")
```

### GO Enrichment Summary
**Upregulated Genes Enrichment**
```{r echo=FALSE}
if (!is.null(ego_up) && nrow(ego_up@result) > 0) {
  top_up <- ego_up@result[1:5, c("Description", "p.adjust")]
  cat("Top enriched biological processes (Upregulated genes):\n")
  print(top_up, row.names = FALSE)
} else {
  cat("No significant GO terms enriched in upregulated genes.\n")
}

cat("\n")
```
**Downregulated Genes Enrichment (relaxed threshold)**
```{r echo=FALSE, warning=FALSE}
if (!is.null(ego_down) && nrow(ego_down@result) > 0) {
  top_down <- ego_down@result[1:5, c("Description", "p.adjust")]
  cat("Top enriched biological processes (Downregulated genes, relaxed p-value < 0.1):\n")
  print(top_down, row.names = FALSE)
} else {
  cat("No significant GO terms enriched in downregulated genes (relaxed threshold).\n")
}
```

## Biological Relevance and Pathway Enrichment

**The analysis of differentially expressed genes identified through this heatmap reveals important biological insights:**

### Upregulated Processes in mTLE

The significantly upregulated genes (22 genes with FDR < 0.05 & log2FC > 1) are predominantly associated with RNA processing and splicing-related pathways. The most enriched biological processes include:

Formation of quadruple SL/U4/U5/U6 snRNP (p.adjust = 7.27e-10)

mRNA trans splicing via spliceosome (p.adjust = 7.27e-10)

Spliceosomal tri-snRNP complex assembly (p.adjust = 1.02e-09)

Spliceosomal snRNP assembly (p.adjust = 1.15e-08)

This strong enrichment suggests that aberrant RNA processing and splicing mechanisms may play a significant role in mTLE pathology, potentially affecting the expression and function of multiple downstream targets.

### Downregulated Processes in mTLE

The downregulated genes (11 genes with FDR < 0.05 & log2FC < -1), analyzed with a relaxed significance threshold (p < 0.1), are associated with:

Vesicle fusion (p.adjust = 0.069)

Organelle membrane fusion (p.adjust = 0.069)

Positive regulation of endocytosis (p.adjust = 0.069)

Membrane fusion (p.adjust = 0.069)

# 9. Supplementary Data: Annotated Differentially Expressed Genes in mTLE Neocortex

```{r include=FALSE}
library(dplyr)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(knitr)

# Ensure your results data frame has gene IDs as a column named 'gene'
# res_df <- as.data.frame(res_neocortex)
# res_df$gene <- rownames(res_df)

# Annotate with gene symbols and gene names
gene_annot <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = res_df$gene,
  keytype = "ENSEMBL",
  columns = c("SYMBOL", "GENENAME")
)

# Merge annotation with your results
res_df_annot <- left_join(res_df, gene_annot, by = c("gene" = "ENSEMBL"))
```
```{r include=FALSE}
# Upregulated genes
sig_up_df <- res_df_annot %>%
  dplyr::filter(padj < 0.05, log2FoldChange > 1) %>%
  dplyr::select(gene, SYMBOL, GENENAME, baseMean, log2FoldChange, padj) %>%
  dplyr::arrange(padj)

# Downregulated genes
sig_down_df <- res_df_annot %>%
  dplyr::filter(padj < 0.05, log2FoldChange < -1) %>%
  dplyr::select(gene, SYMBOL, GENENAME, baseMean, log2FoldChange, padj) %>%
  dplyr::arrange(padj)
```
```{r echo=FALSE}
knitr::kable(sig_up_df, caption = "Significantly Upregulated Genes in mTLE Neocortex")
knitr::kable(sig_down_df, caption = "Significantly Downregulated Genes in mTLE Neocortex")
```

# 10. Data and Code Availability

**All analysis code, reproducible workflow and processed data are available on [GitHub](https://github.com/zeinmursal/mTLE_DGE_Analysis). Raw data is accessible via GEO (GSE134697).**